---
title: "Pset0: Writing more efficient R code for data manipulation"
output: html_document
---

Instructions: 
Please submit your solutions as an exported `.pdf` via `Knit`. 

The purpose of this problem set is to practice writing more efficient code for data manipulation. You will be presented with a series of code chunks that I'm ashamed to have written! Using a combination of vectorized functions and more efficient handling of memory, try to see how fast you can make it run. Your solutions must be faster than implemented functions. Be creative!

Here's the dataset we will be using: a random sample of 5,000 tweets related to the 2014 EP elections in the UK. To make it a bit more challenging, we'll multiply the size of the dataset by 5. We will also make our life easier by removing observations with missing values.

```{r}
d <- read.csv("data/UK-tweets.csv", stringsAsFactors=F)
d <- rbind(d, d, d, d, d)
d <- na.omit(d)
```

## 1) Count how many tweets are classified as impolite (3 points):

Inefficient code

```{r}
func1 <- function(){
  
  count <- 0
  
  for (i in 1:nrow(d)){
    if (d$polite[i]=="impolite"){
      count <- count + 1
    }
  }
  return(count)
}

system.time(count <- func1())

```

Your solution here (remember to test time with system.time()):

```{r}


```

## 2) Count how many tweets mention Brexit (3 points):

Inefficient code

```{r}
func1 <- function(){
  brexit_tweets <- c()
  for (i in 1:nrow(d)){
    if (grepl("brexit", d$text[i], ignore.case=TRUE)==TRUE){
      brexit_tweets <- c(brexit_tweets, i)
    }
  }
  return(length(brexit_tweets))
}
system.time(count <- func1())
```

Your solution that improves test time from the above:

```{r}


```


## 3) Take a random sample of 200 tweets that are classified as "impolite" (4 points):

```{r}
func1 <- function(){
  
  # sample random uniform number
  d$random <- runif(nrow(d))
  # sort by this random number
  dd <- d[order(d$random),]
  # loop over rows using 'while'
  smp <- data.frame()
  i <- 1
  while (nrow(smp)<200){
    if (d$polite[i]=="impolite"){
      smp <- rbind(smp, d[i,])
    }
    i <- i + 1
  }
  return(smp)

}

system.time(smp <- func1())

```

Your solution here (show the efficiency gain):

```{r}


```

