---
title: "Efficient programming with R"
---
Original author: Pablo Barbera (with changes)

```{r}
# run only once
# install.packages("readr")
```

### Vectorization of R code

Efficient programming with R for data analysis requires taking advantage of **vectorized operations**.  

Let's create the following two vectors with random values sampled from a uniform distribution:

```{r}
set.seed(777)
n <- 1000000
x <- runif(n)
y <- runif(n)
```

Which of the following chunks of code will run faster?

```{r}
loop <- function(){
  z <- rep(NA, n)
  for (i in 1:n){
    z[i] <- x[i] + y[i]
  }
  return(z)
}
```

```{r}
sapply_code <- function(){
  z <- sapply(1:n, function(i) x[i] + y[i])
  return(z)
}
```

```{r}
vectorized <- function(){
  z <- x + y
  return(z)
}

```

Let's check with system.time:

```{r}
system.time(loop())
system.time(sapply_code())
system.time(vectorized())
```

There are many other cases where vectorized functions can dramatically speed up our code. The first example shows the power of `ifelse` and the `[` (replacement) function.

```{r}

# suppose we generate n samples, each sample takes a value from 1 to 10, indicating their mood
set.seed(123)
n <- 10000000
vals <- sample(1:10, n, replace=TRUE)

# vectorized code
group_vals <- function(){
  groups <- rep(NA, n)
  groups[vals>5] <- "happy"
  groups[vals<5] <- "sad"
  return(groups)
}

# non-vectorized code
loop_vals <- function(){
  groups <- rep(NA, n)
  for (i in 1:n){
    if (vals[i]>5) groups[i] <- "left"
    if (vals[i]>5) groups[i] <- "right"
  }
}

# benchmarks
system.time(group_vals())
system.time(loop_vals())
```

This second example shows how matrix operations are vectorized too:

```{r}
set.seed(777)
n <- 1000000
df <- data.frame(grade1=runif(n, 0, 100),
                 grade2=runif(n, 0, 100),
                 grade3=runif(n, 0, 100),
                 grade4=runif(n, 0, 100),
                 grade5=runif(n, 0, 100))

system.time(total1 <- df$grade1 + df$grade2 + df$grade3 + df$grade4 + df$grade5)
system.time(total2 <- rowSums(df))
total3 <- rep(NA, n)
system.time(for (i in 1:n){
  total3[i] <- df$grade1[i] + df$grade2[i] + df$grade3[i] + df$grade4[i] + df$grade5[i]
  }
)

# why is rowSums slower than + ?

df <- as.matrix(df)

system.time(total1 <- df[,1] + df[,2] + df[,3] + df[,4] + df[,5])
system.time(total2 <- rowSums(df))
```

The other important way to write more efficient code is to avoid memory copy:

```{r}
set.seed(123)
n <- 10000

# In func1, each call to rnd <- c(rnd, rnorm(n=1)) creates a new vector of length i and copies all i-1 existing elements into it, then adds the new one

func1 <- function(){
  rnd <- c()
  for (i in 1:n){
    rnd <- c(rnd, rnorm(n=1))
  }
  return(rnd)
}

func2 <- function(){
  rnd <- rep(NA, n)
  for (i in 1:n){
    rnd[i] <- rnorm(n=1)
  }
  return(rnd)
}

system.time(func1())
system.time(func2())

# of course, a WAY more efficient way to run this, with vectorized function...

func3 <- function(){
  rnd <- rnorm(n=n)
  return(rnd)
}

system.time(func3())

```

Memory copy applies as well to data frames. A very frequent scenario looks something like the following. You can avoid it by storing results in a list, and then using `do.call(rbind, LIST)` to convert back to a single data frame.


```{r}
set.seed(123)
n <- 200

## using rbind
func1 <- function(){
  df <- data.frame()
  for (i in 1:n){
      df <- rbind(df,  data.frame(
        age = sample(18:100, n, replace=TRUE),
        state = sample(state.abb, n, replace=TRUE),
        party = sample(c("R", "D", "I"), n, replace=TRUE)
      ) )
  }
  return(df)
}

## using lists to avoid memory copy
init <- Sys.time()

func2 <- function(){
  df.list <- list()
  for (i in 1:n){
      df.list[[i]] <- data.frame(
        age = sample(18:100, n, replace=TRUE),
        state = sample(state.abb, n, replace=TRUE),
        party = sample(c("R", "D", "I"), n, replace=TRUE)
      )
  }
  df <- do.call(rbind, df.list)
  return(df)
}


system.time(func1())
system.time(func2())
```

Finding slow spots in your code using `Rprof`

```{r}
Rprof() # start monitoring
invisible(func1()) # run your code (but don't show it!)
Rprof(NULL) # stop
summaryRprof() # see results

```

How to read large files faster? There are different packages to help with this; and also being specific about the data you're reading will help. `readr` is a good option:

```{r}
system.time(d <- read.csv("data/UK-tweets.csv"))
system.time(d <- read.csv("data/UK-tweets.csv", stringsAsFactors=F))
library(readr)
system.time(d <- read_csv("data/UK-tweets.csv"))

```

Ok! You are now done with efficient programming. As you might have noticed, some basics we do in COMP131 are not necessarily the most efficient approach to data manipulation in R. R is primarily written in C, and built around **vectors** as the basic units of computation. The `apply family`, including `apply`, `sapply`, `lapply` are all loops underneath. In future R programming, try looking for **vectorized alternatives** to loops and the `apply` family, if you can!   
